A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-12-12 discussion.tex
final edits

Conversely, we were surprised to find that Ruxanne reported no patterns associated with certain Rust-related features. For instance, we did not observe any patterns associated with the Rust \verb+move+ operator, which has been known to plague many beginning Rust programmers. \verb+move+ is used to transfer the ownership of values to a closure. In some cases (e.g. multithreading), not moving the ownership could create invalid references. Rust compiler checks prevent such errors. Our speculation is that, since Rust enforces strict compiler checks, programmers usually get many compile-time errors, but they only push their code once it is free of compile errors. We thus would not observe patterns which do not make it to the repository.

@@

For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere in the code. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size.

>>>>

Conversely, we were surprised to find that Ruxanne reported no patterns associated with certain Rust-related features. For instance, we did not observe any patterns associated with the Rust \verb+move+ operator, which has been known to plague many beginning Rust programmers. \verb+move+ is used to transfer the ownership of values to a closure. In some cases (e.g. multithreading), not moving the ownership could create invalid references. Rust compiler checks prevent such errors. Our speculation is that, since Rust enforces strict compiler checks, programmers usually get many compile-time errors, but they only push their code once it is free of compile errors. We would not observe patterns which do not make it to the repository.

@@

For instance, as we have discussed, for the common pattern of \textit{Modifying the attributes of structs}, a code linter tool could search for correct attribute lists within the codebase~\citep{forrest2009genetic}, leveraging the assumption that proper attribute usage can be found elsewhere in the code. Also, the linter can analyze the usages of the struct to determine the necessity of adding the attribute. Assuming that attributes cause additional code to be generated, removing unnecessary attributes not only optimizes the codebase but also eliminates unnecessary overhead and reduces binary size.

### 2023-12-12 introduction.tex
final edits

Rust is a relatively new programming language~\citep{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from other systems programming languages (e.g. C/C++) is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties had to be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer from the responsibility to ensure certain safety-related properties~\citep{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\citep{dropbox}.

@@

In this work, with the help of our tool, Ruxanne, we empirically analyze the most popular Rust projects in GitHub. Our goal is to discover common bug patterns that afflict these projects. Although we are motivated by a desire to develop automated program repair tools, we believe these insights can also be of broader service. Automated program repair~\citep{goues2019automated,liu2018survey} tools try to find bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost~\citep{le2012systematic}. A set of common bug fix patterns can be a useful input to an APR system: it tells how to fix recurring buggy structures, thus reducing the search space for patch generation~\citep{jeffrey2009bugfix}. 
Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns in Rust programs.

>>>>

Rust is a relatively new programming language~\citep{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from other systems programming languages (e.g. C/C++) is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties must be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer from the responsibility to ensure certain safety-related properties~\citep{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\citep{dropbox}.

@@

In this work, with the help of our tool, Ruxanne, we empirically analyze the most popular Rust projects in GitHub. Our goal is to discover common bug patterns that afflict these projects. Although we are motivated by a desire to develop automated program repair tools, we believe these insights can also be of broader service. Automated program repair (APR)~\citep{goues2019automated,liu2018survey} tools try to find bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost~\citep{le2012systematic}. A set of common bug fix patterns can be a useful input to an APR system: it tells how to fix recurring buggy structures, thus reducing the search space for patch generation~\citep{jeffrey2009bugfix}. 
Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns in Rust programs.

### 2023-12-12 main.tex
capitalization
final edits

Yet, to our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. Among the general patterns, the most frequently observed pattern is when the user either adds or removes struct fields. In the case of borrow-checker-related patterns, the most common pattern we encountered is when the user removes a clone() call. We describe all detected patterns and their implications to automated program repair.
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

>>>>

Yet, to our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in GitHub to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we identified 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. Among the general patterns, the most frequently observed pattern is when the user either adds or removes struct fields. In the case of borrow-checker-related patterns, the most common pattern we encountered is when the user removes a clone() call. We describe all detected patterns and their implications to automated program repair.
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

### 2023-12-12 methodology.tex
final edits

In our example, we add a \verb+&+ borrow operator to the variable \verb+tail+, which gives rise to an ASTDiff path introducing a new parent node (\verb+ExprReference+) for \verb+tail+. We also add a call to \verb+clone()+. This addition gives rise to two ASTDiff paths: one showing that the \verb+head+ node now belongs to an \verb+ExprMethodCall+, and one indicating the new sibling node for \verb+head+ representing the \verb+clone()+. (Among the terminals, we only keep the distinguished terminal \verb+clone+ in our paths, as it is a BC-related keyword.) One path for the \verb+&+ borrow operator and two for the call to \verb+clone()+ result in three paths in the ASTDiff, even if it only looks like we had made two modifications.

@@

For each cluster, we read the code and the bug reports of the datapoints (if they existed). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging, and removed the datapoints that were not either bug-fix or fix-inducing. We considered these candidates and carried out appropriate ones. Finally, we selected 20 of the remaining clusters---the ones that manifested more-specific changes. Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns} present our clusters.

>>>>

In our example, the change adds a \verb+&+ borrow operator to the variable \verb+tail+, which gives rise to an ASTDiff path introducing a new parent node (\verb+ExprReference+) for \verb+tail+. The change also adds a call to \verb+clone()+. This addition gives rise to two ASTDiff paths: one showing that the \verb+head+ node now belongs to an \verb+ExprMethodCall+, and one indicating the new sibling node for \verb+head+ representing the \verb+clone()+. (Among the terminals, we only keep the distinguished terminal \verb+clone+ in our paths, as it is a BC-related keyword.) One path for the \verb+&+ borrow operator and two for the call to \verb+clone()+ result in three paths in the ASTDiff, even if the change only appears to include two modifications.

@@

For each cluster, we read the code and the bug reports of the datapoints (if available). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging, and removed the datapoints that were not either bug-fix or fix-inducing. We re-considered merging candidates and carried out appropriate merges. Finally, we selected 20 of the remaining clusters---the ones that manifested more-specific changes. Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns} present our clusters.

### 2023-12-12 related.tex
final edits

\cite{pan2009toward} used a bug fix pattern extractor tool to automatically parse and detect bug patterns within bug fix hunks with fewer than seven statements. They ignored larger bug fix hunks---such hunks tended to exhibit random changes rather than having patterns that they could derive meaning from. Though they analyzed Java projects, their reported patterns were not specific to Java, in the same way that our general patterns are not specific to Rust. \cite{martinez2015mining,martinez2012mining} exploited the frequency of observed patterns to introduce a heuristic patch searching method. In their empirical evaluation, they concluded that choosing repair actions probabilistically (weighted by frequency) can help with reducing the search space, hence creating a more effective repair tool.

@@

We would characterize our work as the Rust version of what was done in~\cite{hanam2016discovering} and~\cite{yang2022mining} for JavaScript and Python, respectively. Like~\cite{hanam2016discovering}, we used DBSCAN to cluster fixes. Moreover, similar to~\cite{yang2022mining}, we introduced patterns in two different categories. In our case, there were general and language-specific patterns. The Rust language-specific patterns were related to the borrow checker (BC patterns). However, unlike our paper, \cite{yang2022mining} needed to do a much more elaborate manual analysis, as they had only collected general information about the single hunk changes (e.g. the number of variables, or arguments).

>>>>

\cite{pan2009toward} used a bug fix pattern extractor tool to automatically parse and detect bug patterns within bug fix hunks with fewer than seven statements. They ignored larger bug fix hunks---they claimed that such hunks tended to exhibit random changes rather than having patterns that they could derive meaning from. Though they analyzed Java projects, their reported patterns were not specific to Java, in the same way that our general patterns are not specific to Rust. \cite{martinez2015mining,martinez2012mining} exploited the frequency of observed patterns to introduce a heuristic patch searching method. In their empirical evaluation, they concluded that choosing repair actions probabilistically (weighted by frequency) can help with reducing the search space, hence creating a more effective repair tool.

@@

We would characterize our work as the Rust version of what was done in~\cite{hanam2016discovering} and~\cite{yang2022mining} for JavaScript and Python, respectively, but with additional infrastructure to create a representation of the Rust code using fixed-size vectors. Like~\cite{hanam2016discovering}, we used DBSCAN to cluster fixes. Moreover, similar to~\cite{yang2022mining}, we introduced patterns in two different categories. In our case, there were general and language-specific patterns. The Rust language-specific patterns were related to the borrow checker (BC patterns). However, unlike our paper, \cite{yang2022mining} needed to do a much more elaborate manual analysis, as they had only collected general information about the single hunk changes (e.g. the number of variables, or arguments).

### 2023-12-12 results.tex
final edits

Figure~\ref{fig:rq1} shows the final results. Although we found that our results were acceptable---the result reported by our embedding was top-4 as ranked by humans 88\% of the time---we investigated the other 12\% as well. We found that our embedding tended to fail in change patterns where the change involves many elements (e.g. adding a whole new \verb+impl+ block and implementing new functions in it). Such failing change patterns tend to apply across projects less often, as having many changed elements make it easier for datapoints to be dissimilar.

@@

\textbf{What are the general fix patterns in Rust, and how often do they apply?} 
We collected general fixes, asked DBSCAN to create clusters, and manually analyzed them. In this subsection, we will discuss each pattern, as well as one aspect of its value---its actionability. By actionability, we mean the possible usages of the pattern for bug repair. Other researchers can also use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns; they are not limited to our suggested actionability.  Section~\ref{subsec:patterns} contains a global discussion of our patterns and their value.

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This concept more-or-less corresponds to interfaces in other languages, although the mapping is not exact (e.g. it is possible to implement traits for types that one did not initially define). This change pattern relates to modifications in trait bounds, which can happen as part of bug fixes or for refactoring. \\

@@

In Rust, the match control flow construct is used for comparing a value against multiple patterns, and executing the code associated with the first pattern that matches. A pattern and its associated code are called a match arm. Match statements can be similar to switch statements in other languages, like C, although match statements in Rust are able to match more complicated patterns.\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We applied DBSCAN with parameters $\epsilon=0.001$ and $Z=5$, resulting in 102 clusters. After manual analysis (Section~\ref{sec:manual_analysis_cluster_selection}), we ended up with 8 borrow-checker related fix patterns. Table~\ref{table:bc} describes BC-related patterns, analogously to Table~\ref{table:general} for general patterns. Since the Borrow-Checker is unique to Rust, patterns in this section generally do not have direct counterparts in other languages. We next provide a detailed description of each of these patterns.

@@

\noindent\textbf{Description:} The developer drops the mut keyword before a variable, as they do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed by it. Such a change may help future-proof the code against unintended future changes.

@@

\begin{lstlisting}[language=Rust, style=colouredRust]
pub fn expand_delimited_square(
<@\textcolor{red}{--}@>  children: &Vec<TokenNode>,
<@\textcolor{green}{++}@>  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent\textbf{Description:} In a struct field definition, or in the types of a function's formal parameters, type \verb+&Vec<T>+ changes to slice \verb+&[T]+. Clippy can detect the opportunity to apply this pattern. This simplifies the code; slice types \verb+&[T]+ or \verb+&str+ are sufficient for most use cases.

>>>>

Figure~\ref{fig:rq1} shows the final results. Although we judged that our results were acceptable---the result reported by our embedding was top-4 as ranked by humans 88\% of the time---we investigated the other 12\% as well. We found that our embedding tended to fail in change patterns where the change involves many elements (e.g. adding a whole new \verb+impl+ block and implementing new functions in it). Such failing change patterns tend to apply across projects less often, as having many changed elements make it easier for datapoints to be dissimilar.

@@

\textbf{What are the general fix patterns in Rust, and how often do they apply?} 
We collected general fixes, asked DBSCAN to create clusters, and manually analyzed them. In this subsection, we will discuss each pattern, as well as an important aspect of its value: its actionability. By actionability, we mean the possible usages of the pattern for bug repair. Other researchers can also use the empirical pattern frequency numbers that we report to build their own program repair tools, prioritizing more-frequently-repaired patterns; they are not limited to our suggested actionability.  Section~\ref{subsec:patterns} contains a global discussion of our patterns and their value.

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This concept more-or-less corresponds to interfaces in other languages, although the mapping is not exact (e.g. it is possible to implement traits for others' types). This change pattern relates to modifications in trait bounds, which can happen as part of bug fixes or for refactoring. \\

@@

In Rust, the match control flow construct is used for comparing a value against multiple patterns, and executing the code associated with the first pattern that matches. A pattern and its associated code are called a match arm. Match statements are similar to switch statements in other languages, like C, although Rust match statements are able to match more complicated patterns.\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We then applied DBSCAN with parameters $\epsilon=0.001$ and $Z=5$, resulting in 102 clusters. After manual analysis (Section~\ref{sec:manual_analysis_cluster_selection}), we ended up with 8 borrow-checker related fix patterns. Table~\ref{table:bc} describes BC-related patterns, analogously to Table~\ref{table:general} for general patterns. Since the Borrow-Checker is unique to Rust, patterns in this section generally do not have direct counterparts in other languages. We next provide a detailed description of each of these patterns.

@@

\noindent\textbf{Description:} The developer drops the mut keyword before a variable, as they do not mutate the variable (which is statically checkable). This redundant mutability is reported by Clippy and can also be removed by it. Such a change may help future-proof the code against unintended future changes: mutability must be explicitly added back before the variable can be mutated.

@@

\begin{lstlisting}[language=Rust, style=colouredRust]
pub fn expand_delimited_square(
<@\textcolor{red}{--}@>  children: &Vec<TokenNode>,
<@\textcolor{green}{++}@>  children: &[TokenNode],
) -> Result<hir::Expression, ParseError> {
    // body
}
\end{lstlisting}
\noindent\textbf{Description:} In a struct field definition, or in the types of a function's formal parameters, type \verb+&Vec<T>+ changes to slice \verb+&[T]+. Clippy can detect opportunities to apply this pattern. This simplifies the code; slice types \verb+&[T]+ or \verb+&str+ are sufficient for most use cases.
